---
title: "Assignment 1"
output:
  pdf_document:
    latex_engine: pdflatex
    extra_dependencies: ["float"]
classoption: a4paper
geometry: "left=1cm, right=1cm, top=2cm, bottom=2cm"
fontsize: 10pt
fontfamily: mathptmx
header-includes:
  - \usepackage{wrapfig}
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

# Set the font size for R code
knitr::opts_chunk$set(size = "small")

```


# Read .wea files

Check the common structure of all .wea files across
```{r}
# To use list.files function in R to find all .wea files within all subfolders, you can set the recursive argument to TRUE
weaFiles=list.files(getwd(), pattern = ".wea",recursive = TRUE, full.names = TRUE)
```


```{r}
# read the first few lines of all files
tmp = lapply(weaFiles, readLines, n = 10)
```

We can visually inspect these to see they have the same structure, but different values
that are site-specific.

We can programmatically check the first line starts with place, the second with latitude, etc.
```{r}
expect = c("place", "latitude", "longitude", "time_zone", "site_elevation", 
           "weather_data_file_units")
# sapply, lapply, all, substr
all(lapply(tmp, function(x) all(substr(x[1:6],1,nchar(expect))==expect) ))
```
This is TRUE so we have the same structure (and in the same order) in each file.

 

The data from each file are straightforward to read by skipping the first 6 lines
and noting there are no columb names

```{r}
# first sample lines to read in
ll= lapply(weaFiles, readLines)
ll2=lapply(ll,function(x) x[7:length(x)])
ll3=lapply(ll2, sample, 10)

ll4= lapply(ll3, function(x){
  con=textConnection(x)
  read.table(con, sep = " ", header = FALSE)
})

# check common structure
sapply(ll4,nrow)
sapply(ll4,ncol)
sapply(ll4,str)

# read in all data
dd = lapply(weaFiles, read.table, sep = " ", header = FALSE, skip = 6)
```
We do a quick sanity check
```{r}
sapply(dd, nrow)
```
They each have 8760 rows.

Are they all the same
```{r}
identical(dd[[1]], dd[[2]]) 
```
 
To examine the values, we will stack these data frames along with the location
```{r}
?do.call
# In R, do.call() is a function that allows you to call a function with a list of arguments. It is often used in situations where the arguments to a function are stored in a list or other data structure.
wea = do.call(rbind, dd)
# basename removes all of the path up to and including the last path separator
wea$location = rep(basename(weaFiles), sapply(dd, nrow))
```

Just to check we didn't mess anything up, let's count the number of rows for the different locations we have:
```{r}
table(wea$location)
str(wea)
```
Make some simple vis
```{r,fig.width=5}
library(ggplot2)
library(dplyr)

wea%>%ggplot()+
  geom_density(aes(V4,col=location, group=location),alpha=0.5)

wea%>%ggplot()+
  geom_density(aes(V5,col=location, group=location),alpha=0.5)

#save(wea, file = "wea.Rdata")
```



We define our function as
```{r}
readWEA =function(file)read.table(file, sep = " ", header = FALSE, skip = 6)
```

Finally, we read the 5 WEA files and check the results for basic structure:
```{r}
wea = lapply(weaFiles, readWEA)
sapply(wea, class)
sapply(wea, dim)
```


# Reading a .clm File

The top of a .clm file looks something like
```
*CLIMATE
# ascii climate file from USA_Bodega.Bay.CG.Light.Station EPW file,
# defined in: .\EPW_Output\USA_CA_Bodega.Bay.CG.Light.Station.724995_TMYx.clm
# col 1: Diffuse solar on the horizontal (W/m**2)
# col 2: External dry bulb temperature   (Tenths DEG.C)
# col 3: Direct normal solar intensity   (W/m**2)
# col 4: Prevailing wind speed           (Tenths m/s)
# col 5: Wind direction     (clockwise deg from north)
# col 6: Relative humidity               (Percent)
Bodega.Bay.CG.Light.Station - USA     # site name
 2021,38.31,-3.05,0   # year, latitude, long diff, direct normal rad flag
 1,365    # period (julian days)
* day  1 month  1
 0,75,0,11,329,92
 0,72,0,20,338,91
 0,75,0,13,341,93
 0,61,0,13,338,95
```
The lines col 1, col 2, ... provide the descriptions of the variables.


While we also want the metadata, we'll focus on the values.
These are the rows after the 
```
* day <num> month <num>
```
lines. We only see one of these but there are more.

```{r}
clmFiles=list.files(getwd(), pattern = ".clm",recursive = TRUE, full.names = TRUE)
# read the first few lines of all files
tmp = lapply(clmFiles, readLines, n = 15)
```
```{r}
# check expected pattern for the first 9 lines
expect=c("*CLIMATE","# ascii climate file from","# defined in","# col 1","# col 2","# col 3","# col 4","# col 5","# col 6")

ll=lapply(clmFiles, readLines, n=9)
sapply(ll, function(x) all(substr(x, 1, nchar(expect))==expect))

library(stringr)
# check expected pattern for the 10-12
expect=c("# site name",
         "# year, latitude, long diff, direct normal rad flag",
         "# period (julian days)")
ll=lapply(clmFiles, readLines, n=12)
sapply(ll, function(x) all(str_sub(x[10:12], -nchar(expect), -1)==expect))
 
```

```{r}
dd=lapply(clmFiles, read.table, sep=",", skip=13, header = FALSE)
?read.table
```
  
We can also filter the values out in the call to read.table.
We can change what read.table considers a comment character to *
```{r}
dd=lapply(clmFiles, read.table, sep=",", skip=13, header = FALSE, comment.char = "*")
z2 = read.csv(f, skip = 13, header = FALSE, comment.char = "*")
sapply(dd, class)
sapply(dd, str)
head(dd, 33)
```


While these two approaches allow us to obtain a data.frame,
we have lost the day and month information.
We need to add two columns

There are various approaches we could use, but each relies
on identifying the row identifying the start of a new month.

```{r}
ll=lapply(clmFiles, readLines)
ll=ll[[1]]
# remove the first 13 lines
ll = ll[-(1:12)]

# indicator for day/month lines
isDay=lapply(ll,str_sub, 1,5)=="* day"
str(isDay)

# split each dataset into multiple blocks by * day
groups=cumsum(isDay)
blocks=split(ll,groups)
subdf= lapply(blocks, function(x) read.table(textConnection(x[-1]),header=FALSE,sep=","))
dd=do.call(rbind, subdf)
```

 
We can add the day and month to the entire data.frame, but it may be simpler to do this when
creating each sub-data.frame, i.e., when processing a block.
The first line in each block is of the form
``` 
"* day 25 month 12"
```
We'll separate this by space and exract the third and fifth elements.
Then we can repeat these
```{r}
procCLMBlock=function(x){
  #x=blocks[[1]]
  df=read.table(textConnection(x[-1]),header=FALSE,sep=",")
  dayline=str_split(x[1], " ")[[1]]
  df$day=dayline[3]
  df$month=dayline[5]
  return(df)
}
subdf=lapply(blocks,procCLMBlock)
dd=do.call(rbind, subdf)
# check NA 
head(dd)
```
```{r}
month.dfs = lapply(blocks, procCLMBlock)
```
But this doesn't quite work.
We get warnings

```{r}
# check common structure
checkerror=function(x){
  #x=blocks[[1]]
  df=read.table(textConnection(x[-1]),header=FALSE,sep=",")
  dayline=str_split(x[1], " ")[[1]]
  return(dayline)
}
tmp=lapply(blocks,checkerror)
sapply(tmp,length)
```
 
Here is our problem - the third element is "" not the "2".
Similarly "month" and "1" are separated by "".
This is presumably (assumption) because these are single digit numbers.
When we get to a day with 2 digits, this won't happen for the day, and
similarly for the month.

We can fix this by using a regular expression in the call to `strsplit()`,
but we haven't covered these at this point in the course.
So we just remove "" entries and then the third and fifth elements will be as we expect.
```{r}
procCLMBlock=function(x){
  #x=blocks[[1]]
  df=read.table(textConnection(x[-1]),header=FALSE,sep=",")
  dayline=str_split(x[1], "[\\s|\"]+")[[1]] # els = els[els != ""]
  df$day=as.integer(dayline[3])
  df$month=as.integer(dayline[5])
  return(df)
}
subdf=lapply(blocks,procCLMBlock)
dd=do.call(rbind, subdf)
# check NA 
head(dd) 
# NA for each column
sapply(dd, function(x) all(is.na(x)==FALSE))
```
Good, put into a function calling readCLMBlock inside
```{r}
readCLM=function(f){
  ll = readLines(f)
	ll = ll[-(1:12)]
	isDay=lapply(ll,str_sub, 1,5)=="* day"
  #groups=cumsum(isDay)
  #blocks=split(ll,groups)
	# ?tapply	
	tmp = tapply(ll, cumsum(isDay), procCLMBlock)
	do.call(rbind, tmp)
}
```


```{r}
dd = lapply(clmFiles, readCLM)
names(dd) = basename(clmFiles)
```
We got no error(s) or warnings.
However, we now have to check the values are correct.

We can check the day values range from 1 to 12 at most
and the day values range from 1 to 31.


We can also stack the 5 data.frames together
and compare the distributions for each of the different variables across the different locations.
Before we do this, we should verify 

+ the data frames have the same number of columns
+ the variables in each clm file are the same.

```{r}
sapply(dd, ncol)
table(sapply(dd, ncol))
```
Indeed, they all have 8 columns.


To get the descriptions of the variables, we can read the file
and find the lines starting with `# col`
```{r}
descs= lapply(clmFiles, function(f) grep("# col",readLines(f),value=TRUE))
descs[[1]]
```
  
Are they all the same? We can check they are all the same as the first one:
```{r}
sapply(descs[-1], identical, descs[[1]])
# descs[1]: still list
# descs[[1]]: vector
```
So they are all the same.
 

We can stack the data.frames with 
```{r}
dd2 = do.call(rbind, dd)
dd2$location = rep(names(dd), sapply(dd, nrow))
```
 

We can compare the density of values for V1
```{r}
dd2%>%ggplot()+
  geom_density(aes(x=V1,col=location)) 

dd2%>%ggplot()+
  geom_density(aes(x=V2,col=location))

dd2%>%ggplot()+
  geom_density(aes(x=V3,col=location)) 
dd2%>%ggplot()+
  geom_density(aes(x=V4,col=location)) 

dd2%>%ggplot()+
  geom_density(aes(x=V6,col=location))
```

Of course, the fact that these are similar doesn't mean they are correct.
There could be a bug in our code that makes each location incorrect.

 


We want to put the names of the 6 variables on the data.frame(s).
Again, we'll use regular expressions which we haven't covered yet.

In `descs[[1]]` above, we found the lines starting with `"# col"`.
These look like
```
# col 1: Diffuse solar on the horizontal (W/m**2)
# col 2: External dry bulb temperature   (Tenths DEG.C)
# col 3: Direct normal solar intensity   (W/m**2)
# col 4: Prevailing wind speed           (Tenths m/s)
# col 5: Wind direction     (clockwise deg from north)
# col 6: Relative humidity               (Percent)
```
We want the text after the : but not including the parenthetical part - the units.
```{r}
a = gsub("# col [0-9]: ", "", descs[[1]])
b = gsub("\\(.*", "", a)
names(zz)[1:length(b)] = trimws(b)
```
 


# Read in .stat

```{r}
statFiles=list.files(getwd(), pattern = ".stat",recursive = TRUE, full.names = TRUE)
# read the first few lines of all files
ll= readLines(statFiles[[2]])
 
```



```{r}
# take out all table titles
titles = c("Monthly Statistics for Dry Bulb temperatures",
           "Monthly Statistics for Dew Point temperatures",
           "Average Hourly Statistics for Dry Bulb temperatures",
           "Average Hourly Statistics for Dew Point temperatures",
            "Average Hourly Relative Humidity",
            "Monthly Wind Direction {Interval 11.25 deg from displayed deg)", 
            "Average Hourly Statistics for Direct Normal Solar Radiation",
            "Monthly Statistics for Wind Speed",
            "Average Hourly Statistics for Wind Speed"
)
#str_detect( ll,titles[7]) # grep has issues
 
# check if all table names are right and exist
index=sapply(sapply(titles, function(x) str_detect(ll,fixed(x))),length)

```
All table exists! grep has encoding issue, use str_detect()
 
 


```{r}
# try first table

i = grep(titles[1], ll, value=FALSE, fixed = TRUE)
if(length(i)==0){stop("Can't find table", titles[[1]], " in ", statFiles[[1]] )} #?stop
# subset from the first row of the currect table
ll2 = ll[-(1:i)]
# table always end with summary lines start with -
end=min(which(str_sub(ll2, 1,2 )==" -"|str_sub(ll2, 1,4 )=="   -"))
ll3 = ll2[1:(end - 1L)]
# remove empty lines
ll3 = ll3[ ll3 != ""]
# assign row names by first col
ll4=read.table(textConnection(ll3[-1]),sep="\t",header=F)
# remove all NA columns, first and last
nas=sapply(ll4,function(x) all(is.na(x)))
ll5=ll4[,-which(nas==TRUE)]
# get variables names from the table
cols=trimws(strsplit(ll3[1],"\t")[[1]])
cols=cols[cols!=""]

names(ll5)=c("label", cols )

```


```{r}
# Function: read one table from one file only
readStatFile=function(f,tableTitle){
  # f=statFiles[1]
  #tableTitle=titles[1]
  ll = readLines(f)
  i = min(which(str_detect(ll,fixed(tableTitle))))
  if(length(i)==0){stop("Can't find table ", tableTitle, " in ", f )}  
  # subset from the first row of the currect table
  ll2 = ll[-(1:i)]
  # table always end with summary lines start with -
  end=min(which(str_sub(ll2, 1,2 )==" -"|str_sub(ll2, 1,4 )=="   -"))
  ll3 = ll2[1:(end - 1L)]
  # remove empty lines
  ll3 = ll3[ ll3 != ""]
  # assign row names by first col
  con=textConnection(ll3[-1])
  ll4=read.table(con,sep="\t",header=F)
  on.exit(close(con))
  # remove all NA columns, first and last
  nas=sapply(ll4,function(x) all(is.na(x)))
  ll5=ll4[,-which(nas==TRUE)]
  # get variables names from the table
  cols=trimws(strsplit(ll3[1],"\t")[[1]])
  cols=cols[cols!=""]
  
  names(ll5)=c("label", cols )
  return(ll5)
}
```

```{r}
# Function: return a list of data.frames with names
tmp=lapply(titles,readStatFile, f=statFiles[1])
names(tmp)=titles
tmp
str(tmp)
```

For monthly data:
Verify that the Max Hour and Min Hour are correct. – then omit these rows
• Convert the data so that the rows corresponding to measured variables and dates e.g. Maximum, Minimum, Daily Avg, . . . are columns and the columns corresponding to months are rows.
• Convert the Day:Hour values to a time (POSIXct). Use 2023 as the year.
• Convert the measurements for other variables to numbers.

```{r}
cvtDayHour =function(x, month)
{    
    # x=tmp[[1]]
    # month='Jan'
    str = sprintf("2023/%s/%s:00", month, trimws(x))
    strptime(str, "%y/%b/%d:%H:%M")
    as.POSIXct(strptime(str, "%Y/%b/%d:%H:%M"))
    
}
 
```

```{r}
transMonthly=function(d){
  d=tmp[[1]]
  r.index= str_detect(d[,1],"Day:Hour") # which row corresponds to day hour
  tmp=t(d[!r.index,-1]) # transpose and remove first name col and few rows
  mode(tmp)="numeric"
  d2=as.data.frame(tmp)
  colnames(d2)=d[!r.index, 1]
  
  # add day:hour columns 
  tmp=as.data.frame(t(d[r.index,-1]))
  tmpnames=d[which(r.index)-1L,1]
  tmpnames=paste0(tmpnames, "Time")
  d2[tmpnames]=lapply(tmp, cvtDayHour, names(d)[-1])
  return(d2)
}

transMonthly(tmp[[1]])
```

Apply to all monthly datasets
```{r}
# Function: return a list of data.frames with names
tmp=lapply(titles,readStatFile, f=statFiles[1])
names(tmp)=titles

index=which(str_detect(titles[-6], "Monthly"))
tmp[titles[index]]=lapply(tmp[titles[index]], transMonthly)
tmp
 
```

For the hourly data tables,
• convert each to a data.frame with 3 columns:
– converting the month-hour pairs to rows with the single variable as a column – one column for the month
– one column for the hour 
```{r}
transHourly =function(d, varName = names(d)[1])
{   
  
    # wide -> long
    #d=tmp[[3]]
    # remove extra lines
    r.index= str_detect(d[,1],"Hour")
    d = d[!r.index, ]
    
    vals=as.numeric(unlist(d[,-1]))
    tmp=data.frame(value=vals, month=rep(names(d)[-1],each=dim(d)[1]),
                   hour=rep(trimws(str_sub(d[,1],1,2)),ncol(d)-1))
    
    return(tmp)
}

# Function: return a list of data.frames with names
tmp=lapply(titles,readStatFile, f=statFiles[1])
names(tmp)=titles

index=which(str_detect(titles[-6], "Hourly"))
tmp[titles[index]]=lapply(tmp[titles[index]], transHourly)
tmp 

```

"Monthly Wind Direction {Interval 11.25 deg from displayed deg): Different!




Finally,
• combine the average hourly tables into a single data frame with a column for each variable, i.e., dry bulb temperature, dew, relative radiation, wind speed. Ensure that the rows correspond to the same time, i.e., month, hour and day.
• for each variable, plot the values against hour for each month.

```{r}
# Function: return a list of data.frames with names
tmp=lapply(titles,readStatFile, f=statFiles[1])
names(tmp)=titles

titles=titles[-6]
tmp=tmp[-6]
index=which(str_detect(titles, "Hourly"))
tmp[titles[index]]=lapply(tmp[titles[index]], transHourly)
tmp.hr=tmp[titles[index]]

# first check month and hour are identical
sapply(tmp.hr[-1], function(x) identical(tmp.hr[[1]][c("month","hour")],x[c("month","hour")]))

```
```{r}
cdd=do.call(cbind, tmp.hr)
names(cdd)
keepi=which(str_detect(names(cdd),"value"))
cdd=cdd[names(cdd)[c(keepi,11:12)]]
names(cdd)=c("Dry Bulb temperatures","Dew Point temperatures","Relative Humidity",                    
              "Wind Speed","Month","Hour")

cdd%>%ggplot(aes(x=Hour, y=`Dry Bulb temperatures`,group=Month, col=Month))+
  geom_line()

cdd%>%ggplot(aes(x=Hour, y=`Dew Point temperatures`,group=Month, col=Month))+
  geom_line()

cdd%>%ggplot(aes(x=Hour, y=`Relative Humidity`,group=Month, col=Month))+
  geom_line()

 
 
  
```

